---
title: "Chap3"
author: "Paloma"
date: "11/28/2019"
output: 
  html_document:
    df_printed: paged
    theme: united
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Chapter 3

## 3.1 Intro


## 3.2 First steps

The idea is that we have:
`ggplot(data) + <GEOM_FUNCTION(aes(), other feat>`

```{r out.width=c('50%','50%'), fig.show='hold'}
str(mpg)

dim(mpg)

ggplot(data = mpg,
       aes(hwy, cyl)) +
  geom_point()

ggplot(data = mpg,
       aes(class, drv)) +
  geom_point()

```

We cant see the distribution of observations across categories.

## 3.3 Aesthetic mapping

Key tip:
`group`: does not give a legend

`color`: gives a legend, we need to `set.legend = FALSE` if we don't want it.

`size`: not advised for discrete categories, ideally for weights or a count variable

`shape`: is good for categorical, but it only has 25 built in shapes.

`fill`: can also give color to some geom, but not `geom_plot`. I think it is similar to the shape/color and fill relationship.

```{r out.width=c('50%','50%'), fig.show='hold'}
ggplot(data = mpg,
       aes(hwy, cyl, group = class)) +
  geom_point() +
  labs(title = "using group argument")

ggplot(data = mpg,
       aes(hwy, cyl, color = class)) +
  geom_point() +
    labs(title = "using color argument")

ggplot(data = mpg,
       aes(hwy, cyl, shape = class)) +
  geom_point() +
    labs(title = "using shape argument")

ggplot(data = mpg,
       aes(hwy, cyl, fill = class)) +
  geom_point() +
    labs(title = "using fill argument")

ggplot(data = mpg,
       aes(class, hwy)) +
  geom_boxplot(aes(fill = class)) +
    labs(title = "using fill argument")


```

If we want only one shape or color, that would NOT have to be inside the `aes()`, but as an argument of the `<geom_function()>`

About the `shape`, it goes with `color` and `fill`.

- From 0 to 14 *borders* are determined by `color`

- From 15 to 18 are *filled* by `color`

- From 21 - 24 the *border* is defined by `color` and *filled* 
with `fill`

<div class = "row">
<div class = "col-md-4">
```{r}
ggplot(data = mpg, aes(hwy, cyl)) +
  geom_point(shape = 3, color = "orange", size = 5) +
    labs(title = "using shape argument")
```
</div>

<div class = "col-md-4">
```{r}
ggplot(data = mpg, aes(hwy, cyl)) +
  geom_point(shape = 16, color = "orange", size = 5) +
    labs(title = "using shape argument")
```
</div>

<div class = "col-md-4">
```{r}
ggplot(data = mpg, aes(hwy, cyl)) +
  geom_point(shape = 22, color = "orange", fill = "yellow", size = 5) +
    labs(title = "using shape argument")
```
</div>
</div>

## 3.4 Common problems

Just remeber to use `?function_name`.

## 3.5 Facets

`facet_wrap` = makes rows, and usually goes like `facet_wrap(~ cat_var)`

We can add `nrow` and `ncol`, `switch` changes where the title of each category goes, could be "x", "y" or "both".

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 4, ncol = 2, switch = "y")
```


- `facet_grid` = Makes columns, it usually goes like `facet_grid(.~ cat_var)`

<div class = "row">
<div class = "col-md-6">
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ class)
```
</div>

<div class = "col-md-6">
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ class)
```
</div>
</div>



## 3.6 Geometric objects
## 3.7 Statistical transformations
## 3.8 Position adjustments
## 3.9 Coordinate systems
## 3.10 The layered grammar of graphics



