---
title: "Chapter 5"
author: "Paloma"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_printed: paged
    theme: united
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)

mytable <- function(df){
  kable(df) %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"),
                full_width = F,
                font_size = 12)}
```

# Data transformation
## Intro

Some key concepts

**Different types of variables**

- `int` as integers

- `dbl` stands for doubles, real numbers

- `chr` stands for character vectors, or strings

- `dttm` stands for date-times (a date + a time)

- `lgl` logical, TRUE or FALSE

- `fctr` factors, categorical variables with fixed possible values

- `date` stands for dates

About `dplyr` verbs:

1. First argument is a data frame

2. The subsequent argument describe what to do with the data frame, using the variable names (_without quotes_).

3. The result is a new data frame

## 5.2 Filter rows with `filter`

**Def**: _subset observations based on their values._

**Logical operators**

`x`
```{r}
table2 %>% 
  filter(country == "Brazil") %>% 
  mytable()
```

`y`
```{r}
table2 %>% 
  filter(type == "cases") %>% 
  mytable()
```

`x & y`
```{r}
table2 %>% 
  filter(type == "cases" & country == "Brazil") %>% 
  mytable()
```

`x | y`
```{r}
table2 %>% 
  filter(type == "cases" | country == "Brazil") %>% 
  mytable()
```

`y & !x`
```{r}
table2 %>% 
  filter(type == "cases" & country != "Brazil") %>% 
  mytable()
```

`x & !y`
```{r}
table2 %>% 
  filter(country == "Brazil" & type != "cases") %>% 
  mytable()
```

`xor(x,y)`
```{r}
table2 %>% 
  filter(xor(country == "Brazil", 
             type == "cases")) %>% 
  mytable()
```

`%in%`

```{r}
table2 %>% 
  filter(country %in% c("Brazil", "China")) %>% 
  mytable()
```

`!(x&y) = !x|!y`

```{r}
table2 %>% 
  filter(!(country == "Brazil" & type == "cases"))%>% 
  mytable()

table2 %>% 
  filter(country != "Brazil" | type != "cases")%>% 
  mytable()

```

`!(x|y) = !x & !y`

```{r}
table2 %>% 
  filter(!(country == "Brazil" | type == "cases"))%>% 
  mytable()

table2 %>% 
  filter(country != "Brazil" & type != "cases")%>% 
  mytable()
```

filter() only includes rows where the condition is TRUE; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly:

```{r}
df <- tibble(x = c(1, NA, 3))

filter(df, x > 1)

filter(df, is.na(x) | x > 1)
```

1. Find all flights that

```{r}
library(nycflights13)
dim(flights)
summary(flights) %>% 
  mytable()
```

  1. Had an arrival delay of two or more hours
  
```{r}
 flights %>% 
  filter(arr_delay >= 120) %>% 
  count() %>% 
  mytable()
```

  2. Flew to Houston (IAH or HOU)
  
```{r}
flights %>% 
  filter(dest %in% c("IAH", "HOU")) %>% 
  count(dest) %>% 
  mytable()
```
  
  3. Were operated by United, American, or Delta

```{r}
flights %>% 
  filter(carrier %in% c("UA", "AA", "DL")) %>% 
  count(carrier) %>% 
  mytable()
```

  4. Departed in summer (July, August, and September)
```{r}
flights %>% 
  filter(between(month,7,9)) %>% 
  count(month) %>% 
  mytable()
```
  5. Arrived more than two hours late, but didn’t leave late
```{r}
flights %>% 
  filter(dep_delay <= 0 & arr_delay > 120) %>% 
  count() %>% 
  mytable()
```  
  
  6. Were delayed by at least an hour, but made up over 30 minutes in flight
  
```{r}
flights %>% 
  filter(dep_delay >= 60 & (dep_delay - arr_delay >= 30)) %>% 
  count() %>% 
  mytable()
```  
  7. Departed between midnight and 6am (inclusive)
```{r}
flights %>% 
  filter(dep_time <= 600 | dep_time == 2400) %>% 
  count() %>% 
  mytable()

flights %>%
  filter(dep_time %in% c(1:600, 2400)) %>%
  count() %>% 
  mytable()
```  

3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

```{r}
flights %>% 
count(is.na(dep_time)) %>% 
  mytable()

```

4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing?. 5. Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)

```{r}
NA^0

NA|TRUE

NA|FALSE

NA&TRUE

NA&FALSE

NA*0
```

## 5.3 Arrange rows with `arrange`

1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

```{r}
flights %>% 
  arrange(desc(is.na(dep_time)), dep_time) %>% 
  slice(1, 1000,20000,300000) %>% 
  mytable()
```

2. Sort flights to find the most delayed flights. Find the flights that left earliest.

```{r}
# Most delayed
flights %>% 
  arrange(desc(dep_delay)) %>% 
  slice(1:5) %>% select(carrier, flight, dep_delay) %>% 
  mytable()

# Earliest
flights %>% 
  arrange(dep_delay) %>% 
  slice(1:5) %>% select(carrier, flight, dep_delay) %>% 
  mytable()
```

3. Sort flights to find the fastest flights.

```{r}
flights %>% 
  mutate(mph = distance/(air_time/60)) %>% 
  arrange(desc(mph)) %>% 
  slice(1:5) %>% select(carrier, flight, air_time, distance) %>% 
  mytable()
```


Which flights travelled the longest? Which travelled the shortest?

```{r}
flights %>% 
  select(carrier, flight, distance) %>% 
  arrange(distance) %>% 
  slice(1, n()) %>% 
  mytable()
  
```

## 5.4 Select rows with `select`
## 5.5 Add new variables with `mutate`
## 5.6 Grouped summaries with `summarise`
## 5.7 Grouped mutates (and filters)





