---
title: "Chapter 5"
author: "Paloma"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size = 3)
library(tidyverse)
library(kableExtra)

mytable <- function(df){
  kable(df) %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"),
                full_width = F,
                font_size = 12)}
```

# Data transformation
## Intro

Some key concepts

**Different types of variables**

- `int` as integers

- `dbl` stands for doubles, real numbers

- `chr` stands for character vectors, or strings

- `dttm` stands for date-times (a date + a time)

- `lgl` logical, TRUE or FALSE

- `fctr` factors, categorical variables with fixed possible values

- `date` stands for dates

About `dplyr` verbs:

1. First argument is a data frame

2. The subsequent argument describe what to do with the data frame, using the variable names (_without quotes_).

3. The result is a new data frame

## 5.2 Filter rows with `filter`

**Def**: _subset observations based on their values._

**Logical operators**

`x`
```{r}
table2 %>% 
  filter(country == "Brazil") %>% 
  mytable()
```

`y`
```{r}
table2 %>% 
  filter(type == "cases") %>% 
  mytable()
```

`x & y`
```{r}
table2 %>% 
  filter(type == "cases" & country == "Brazil") %>% 
  mytable()
```

`x | y`
```{r}
table2 %>% 
  filter(type == "cases" | country == "Brazil") %>% 
  mytable()
```

`y & !x`
```{r}
table2 %>% 
  filter(type == "cases" & country != "Brazil") %>% 
  mytable()
```

`x & !y`
```{r}
table2 %>% 
  filter(country == "Brazil" & type != "cases") %>% 
  mytable()
```

`xor(x,y)`
```{r}
table2 %>% 
  filter(xor(country == "Brazil", 
             type == "cases")) %>% 
  mytable()
```

`%in%`

```{r}
table2 %>% 
  filter(country %in% c("Brazil", "China")) %>% 
  mytable()
```

`!(x&y) = !x|!y`

```{r}
table2 %>% 
  filter(!(country == "Brazil" & type == "cases"))%>% 
  mytable()

table2 %>% 
  filter(country != "Brazil" | type != "cases")%>% 
  mytable()

```

`!(x|y) = !x & !y`

```{r}
table2 %>% 
  filter(!(country == "Brazil" | type == "cases"))%>% 
  mytable()

table2 %>% 
  filter(country != "Brazil" & type != "cases")%>% 
  mytable()
```

filter() only includes rows where the condition is TRUE; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly:

```{r}
df <- tibble(x = c(1, NA, 3))

filter(df, x > 1)

filter(df, is.na(x) | x > 1)
```

1. Find all flights that

```{r}
library(nycflights13)
dim(flights)
```

  1. Had an arrival delay of two or more hours
  
```{r}
 flights %>% 
  filter(arr_delay >= 120) %>% 
  count() %>% 
  mytable()
```

  2. Flew to Houston (IAH or HOU)
  
```{r}
flights %>% 
  filter(dest %in% c("IAH", "HOU")) %>% 
  count(dest) %>% 
  mytable()
```
  
  3. Were operated by United, American, or Delta

```{r}
flights %>% 
  filter(carrier %in% c("UA", "AA", "DL")) %>% 
  count(carrier) %>% 
  mytable()
```

  4. Departed in summer (July, August, and September)
```{r}
flights %>% 
  filter(between(month,7,9)) %>% 
  count(month) %>% 
  mytable()
```
  5. Arrived more than two hours late, but didn’t leave late
```{r}
flights %>% 
  filter(dep_delay <= 0 & arr_delay > 120) %>% 
  count() %>% 
  mytable()
```  
  
  6. Were delayed by at least an hour, but made up over 30 minutes in flight
  
```{r}
flights %>% 
  filter(dep_delay >= 60 & (dep_delay - arr_delay >= 30)) %>% 
  count() %>% 
  mytable()
```  
  7. Departed between midnight and 6am (inclusive)
```{r}
flights %>% 
  filter(dep_time <= 600 | dep_time == 2400) %>% 
  count() %>% 
  mytable()

flights %>%
  filter(dep_time %in% c(1:600, 2400)) %>%
  count() %>% 
  mytable()
```  

3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

```{r}
flights %>% 
count(is.na(dep_time)) %>% 
  mytable()

```

4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing?. 5. Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)

```{r}
NA^0

NA|TRUE

NA|FALSE

NA&TRUE

NA&FALSE

NA*0
```

## 5.3 Arrange rows with `arrange`

1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

```{r}
flights %>% 
  arrange(desc(is.na(dep_time)), dep_time) %>% 
  select(1:4) %>% 
  slice(1, 1000,20000,300000) %>% 
  mytable()
```

2. Sort flights to find the most delayed flights. Find the flights that left earliest.

```{r}
# Most delayed
flights %>% 
  arrange(desc(dep_delay)) %>% 
  slice(1:5) %>% select(carrier, flight, dep_delay) %>% 
  mytable()

# Earliest
flights %>% 
  arrange(dep_delay) %>% 
  slice(1:5) %>% select(carrier, flight, dep_delay) %>% 
  mytable()
```

3. Sort flights to find the fastest flights.

```{r}
flights %>% 
  mutate(mph = distance/(air_time/60)) %>% 
  arrange(desc(mph)) %>% 
  slice(1:5) %>% select(carrier, flight, air_time, distance) %>% 
  mytable()
```


Which flights travelled the longest? Which travelled the shortest?

```{r}
flights %>% 
  select(carrier, flight, distance) %>% 
  arrange(distance) %>% 
  slice(1, n()) %>% 
  mytable()
  
```

## 5.4 Select rows with `select`

`select(var, -(varx:varz))`

`num_range("x", 1:3)`: matches x1,x2,x3

`one_of` is old

`contains` ignores case, but has an argument `ignore.case`.

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

flights %>% 
  select(one_of(vars)) %>% 
  colnames()

flights %>% 
  select(contains("TIME")) %>% 
  colnames()


flights %>% 
  select(contains("TIME", ignore.case = FALSE)) %>% 
  colnames()

```
## 5.5 Add new variables with `mutate`

- `transmute` only keeps the new variables.

- Arithmetiic operators also work in conjunction with aggregate functions. For ex. `x/sum(x)`, `y - mean(y)`.

- Modular arithmetics: 

  + `%/%` = integer division

  + `%%` = remainder
  
```{r}
transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100) %>% 
  head() %>% 
  mytable()
```

- `log()` (by default natural logarithms), `log2()` (base 2), `log10()` (base 10)

- `lead()`, `lag()`

- `log1p(x)` computes log(1+x) accurately also for |x| << 1

- `cumsum()`, `cumprod()`, `cummin()`, `cummean()`

```{r}
x <- c(1:10)

cumsum(x)

cummean(x)

cumprod(x)

cummin(x)

cummax(x)
```

- Ranking

```{r}
y <- c(1, 2, 2, NA, 3, 4)
x <- c(5, 1, 3, 2, 2, NA)

min_rank(x)
row_number(x)
min_rank(desc(x))

min_rank(y)
row_number(y)
min_rank(desc(x))

dense_rank(y)

percent_rank(y)

cume_dist(y)

ntile(x, 0.9)
```

**Exercises**

1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

```{r}
flights %>%
  transmute(
    dep_time  = dep_time,
    dep_time_hr = (dep_time %/% 100),    # gets the hours
    dep_time_hr = dep_time_hr * 60,    #converts to minutes
    dep_time_min = (dep_time %% 100),
    dep_time_min = dep_time_min + dep_time_hr
  ) %>%
  mutate(dep_time_min = ifelse(dep_time == 2400, 0, dep_time_min)) %>% 
  head() %>% 
  mytable()
```

2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?

```{r}
flights %>%
  sample_n(10000) %>% 
  select(air_time, arr_time, dep_time) %>%
  mutate(arr_dep = arr_time - dep_time) %>% 
  mutate(equal = air_time == arr_dep) %>% 
  ggplot(aes(air_time, arr_dep)) +
  geom_point()
```

3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

```{r}
flights %>%
  select(dep_time, sched_dep_time, dep_delay) %>% 
  head() %>% 
  mytable()
```

4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().

```{r}
flights %>% 
  select(origin, carrier, dep_delay) %>% 
  # arrange(desc(dep_delay)) %>% 
  mutate(minrank = min_rank(desc(dep_delay)),
         rowrank = row_number(desc(dep_delay)),
         denserank = dense_rank(desc(dep_delay))) %>% 
  arrange(desc(dep_delay)) %>% 
  head() %>% 
  mytable()
```

5. What does 1:3 + 1:10 return? Why?

```{r}
1:3 + 1:10
```

6. What trigonometric functions does R provide?

`sin()`, `cos`, `tan`

## 5.6 Grouped summaries with `summarise`

Useful summarise functions:

- `mean()`, `median()`, `sd()`, `IQR()`, `mad()`

- `min()`, `quantile(x, 0.5)`, `max()`

- `first()`, `nth(x,2)`, `last()`

- `n()`, `n_distinct()`, `sum(!is.na(x))`

- `count()`, `count(x, wt = y)`


```{r}
flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    mean_delay = mean(arr_delay, na.rm = TRUE),
    mean_dist = mean(distance, na.rm = TRUE)) %>%
  filter(mean_dist <3000) %>% 
  ggplot(aes(mean_dist, mean_delay)) +
  geom_point(aes(size = count), alpha = 0.5) +
  geom_smooth()
```

```{r}
flights %>% 
  group_by(tailnum) %>% 
  summarize(
    count = n(),
    mean_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  filter(between(count, 25, 500)) %>% 
  ggplot(aes(count, mean_delay)) +
  geom_point(alpha = 0.5, position = "jitter")
```

```{r}
summary_flights <- flights %>% 
  select(dest, arr_delay) %>% 
  filter(!is.na(arr_delay)) %>% 
  group_by(dest) %>% 
  summarize(mean = mean(arr_delay),
            mean_pos = mean(arr_delay[arr_delay > 0]),
            median = median(arr_delay),
            sd = sd(arr_delay[arr_delay > 0]),
            iqr = IQR(arr_delay),
            mad = mad(arr_delay),
            min = min(arr_delay),
            quantile50 = quantile(arr_delay, 0.5),
            max = max(arr_delay),
            first = first(arr_delay),
            last = last(arr_delay))
summary_flights %>% 
  head() %>% 
  mytable()
```

```{r}
summary_flights %>%
  sample_n(10) %>% 
  ggplot(aes(x = dest, y = median)) +
  geom_bar(aes(fill = dest), stat = "identity") +
  coord_flip() 

```

```{r}
summary_flights %>%
  sample_n(10) %>% 
  ggplot(aes(x = dest, y = median)) +
  geom_col(aes(fill = dest)) +
  coord_flip() 

```

```{r}
summary_flights %>% 
  sample_n(5) %>% 
  ggplot(aes(x = dest, y = mean_pos, color = dest)) +
  geom_pointrange(aes(ymin = min, ymax = max)) +
  coord_flip()

```

```{r}
summary_flights %>% 
  sample_n(5) %>% 
  ggplot(aes(x = dest, y = mean_pos, color = dest)) +
  geom_pointrange(aes(ymin = min, ymax = max)) +
  geom_errorbar(aes(ymin = min, ymax = max)) +
  coord_flip()

```

```{r}
summary_flights %>%
  sample_n(5) %>% 
  ggplot(aes(x = dest, y = mean_pos)) +
  geom_bar(aes(color = dest), fill = NA, stat = "identity", show.legend = FALSE) +
  geom_pointrange(aes(ymin = min, ymax = max, color = dest)) +
  geom_errorbar(aes(ymin = min, ymax = max, color = dest)) +
  geom_hline(yintercept = 0, color = "black", size = 1) +
  coord_flip()

```

**Calculate proportions**

Where do people go?

```{r}
flights %>%
  select(dest) %>% 
  group_by(dest) %>% 
  mutate(n_dest = n()) %>% 
  ungroup() %>% 
  filter(n_dest > 5000) %>% 
  count(dest, name = "total") %>% 
  mutate(prop = round(total/sum(total)*100,2)) %>% 
  mutate(dest = fct_reorder(dest, prop)) %>% 
  ggplot(aes(dest, prop, fill = dest)) +
  geom_col(show.legend = FALSE) +
  geom_label(
    aes(label = prop, y = prop), show.legend = FALSE)
```

**Proportions for a dicotomous cathegory**

```{r}
flights %>% 
  summarise(mean(arr_delay < 0, na.rm = TRUE))
```

```{r}
flights %>%
  mutate(
    arr_cat = case_when(
      arr_delay < 0 ~ "earlier",
      arr_delay == 0 ~ "sharp",
      arr_delay > 0 ~ "late",
      is.na(arr_delay) ~ "cancelled"
    )
  ) %>%
  count(arr_cat, name = "total") %>%
  mutate(prop = total / sum(total),
         total_count = sum(total))

```

```{r}
flights %>%
  sample_n(100) %>% 
  ggplot(aes(dest))+
  geom_bar(aes(y = ..prop.., group = 1))
```

```{r}
flights %>% 
  sample_n(500) %>% 
  count(dest, name = "count") %>% 
  mutate(prop = round(count/sum(count)*100,2),
         dest = fct_reorder(dest, prop)) %>%
  filter(prop > 1) %>% 
  ggplot(aes(dest, prop)) +
  geom_col(aes(fill=dest)) +
  geom_label(aes(label = prop, y = prop)) +
  coord_flip()
```

## 5.7 Grouped mutates (and filters)







