---
title: "Chapter 15"
author: "Paloma"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size = 3)
library(tidyverse)
library(kableExtra)

mytable <- function(df){
  kable(df) %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"),
                full_width = F,
                font_size = 12)}
```

## Factors

Order of levels can be 

- Arbitrary

- Principled: ordinal or ordered by the count

### Relevel by hand()

```{r}
labels <- c("Jan", "Mar", "Apr", "June", "Dec")
x1 <- c("Dec", "Apr", "Jan", "Mar", "June")

x1 <- enframe(x1) %>% 
  mutate(value = fct_relevel(as_factor(value), labels))

levels(x1$value)

x1 <- tibble(
  month = rep(c("Dec", "Apr", "Jan", "June", "Mar"), times = 4),
  count = rnorm(20, mean = 50, sd = 20))

levels(x1$month)

x1 %>% 
ggplot(aes(month, count, fill = month)) +
  geom_boxplot()

x1 %>% 
  mutate(month = fct_relevel(month, labels)) %>%
  ggplot(aes(month, count, fill = month)) +
  geom_boxplot()

```

We can also pull first one cathegory to the front
```{r}
x1 %>% 
  mutate(month = fct_reorder(month, count)) %>%
  mutate(month = fct_relevel(month, c("Dec", "Mar"))) %>%
  ggplot(aes(month, count, fill = month)) +
  geom_boxplot()

```

Or to the back 

```{r}
x1 %>% 
  mutate(month = fct_reorder(month, count)) %>%
  mutate(month = fct_relevel(month, "Dec", after = Inf)) %>%
  ggplot(aes(month, count, fill = month)) +
  geom_boxplot()

```

### Reorder by frequency (and reverse)

```{r}
x1 %>% 
  mutate(month = fct_reorder(month, count)) %>%
  ggplot(aes(month, count, fill = month)) +
  geom_boxplot()

x1 %>% 
  mutate(month = fct_rev(fct_reorder(month, count))) %>%
  ggplot(aes(month, count, fill = month)) +
  geom_boxplot()
```


### Empty categories

Usually they are dropped if the cathegory doesn't have any values, so we need to make it explicit

```{r}
forcats::gss_cat %>% 
  ggplot(aes(race)) +
  geom_bar()

forcats::gss_cat %>% 
  ggplot(aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

**Exercise**

Explore the distribution of rincome (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

```{r}
df <- forcats::gss_cat

df %>% 
  count(rincome) %>% mytable()

df %>%
  group_by(rincome) %>% 
  mutate(count_inc = n()) %>% 
  ungroup() %>% 
  mutate(rincome = fct_reorder(rincome, count_inc)) %>% 
  ggplot(aes(rincome)) +
  geom_bar(aes(fill = rincome), show.legend = FALSE) + 
  coord_flip()

```

What is the most common relig in this survey? What’s the most common partyid?

```{r}
df %>% 
  count(relig, sort = TRUE) %>% 
  slice(1) %>% mytable()

df %>% 
  count(partyid, sort = TRUE) %>% 
  slice(1) %>% mytable()
```

Which relig does denom (denomination) apply to? How can you find out with a table? How can you find out with a visualization?

```{r}
df %>% 
  count(relig, denom, sort = TRUE) %>% 
  ggplot(aes(relig, denom)) +
  geom_tile(aes(fill = n)) +
  theme(axis.text.x = element_text(angle = 90, size = 8),
        axis.text.y = element_text(size = 8))
```

### Reorder2

Reorders the factor by the y values associated with the largest x values. 
```{r}
age_marital <- df %>%
  filter(!is.na(age)) %>%
  count(age, marital) %>%
  group_by(age) %>%
  mutate(prop = n / sum(n))

age_marital %>% 
  ggplot(aes(age, prop, color = marital)) +
  geom_line(size = 1)

age_marital %>% 
  ggplot(aes(age, prop, color = fct_reorder2(marital, age, prop))) +
  geom_line(size = 1)

x1 %>%
  mutate(y = round(rnorm(20, mean = 0, sd = 5), 0)) %>% 
  ggplot(aes(y, count, color = fct_reorder2(month, y, count))) +
  geom_line(size = 1)
```

### In frequence: fct_infreq()

```{r}
df %>% 
  ggplot(aes(fct_infreq(rincome))) +
  geom_bar() +
  theme(
    axis.text.x = element_text(angle = 90))

df %>% 
  ggplot(aes(fct_rev(fct_infreq(rincome)))) +
  geom_bar() +
  theme(
    axis.text.x = element_text(angle = 90))

```

**Exercises**

There are some suspiciously high numbers in tvhours. Is the mean a good summary?
```{r}
df %>% 
  ggplot(aes(tvhours)) +
  geom_histogram()
```

For each factor in gss_cat identify whether the order of the levels is arbitrary or principled.
```{r}
df %>% 
  select_if(is.factor) %>% 
  names()
```

### Recode

Allows to change the value of each level, factors don't not work with `case_when`.

It leaves all unmentioned factors as they are.

```{r,error=TRUE}
df %>% 
  count(partyid) %>% 
  head() %>% mytable()

df %>%
  mutate(partyid = case_when(
    partyid == "No answer" ~ "NoA",
    TRUE ~ partyid)) %>% 
  count(partyid)

df %>%
  mutate(partyid = case_when(
    partyid == "No answer" ~ "NoA",
    TRUE ~ "Others")) %>% 
  count(partyid)
```

```{r}
df %>%
  mutate(partyid = fct_recode(partyid,
    "NoA" = "No answer")) %>% 
  count(partyid) %>% 
  head() %>% mytable


```

### Collapse

```{r}
gss_cat %>% 
count(rincome, sort = TRUE) %>% 
  mutate(rincome = fct_collapse(rincome,
        "Not available" = c("Not applicable",
                            "Refused",
                            "Don't know",
                            "No answer"),
        "$1000 - 7999" = c("Lt $1000",
                           "$1000 to 2999",
                           "$3000 to 3999",
                           "$4000 to 4999",
                           "$5000 to 5999",
                           "$6000 to 6999",
                           "$7000 to 7999"))) %>% 
  mytable()
                           
```

### Lump

```{r}
df %>% 
  mutate(rincome = fct_lump(rincome, n = 5)) %>% 
  count(rincome)
```

**Exercises**

How have the proportions of people identifying as Democrat, Republican, and Independent changed over time?

```{r}
df %>% 
  count(partyid)

df_clean <- df %>% 
  select(year, partyid) %>% 
  mutate(
    partyid = fct_collapse(partyid,
                         republican = c("Strong republican",
                                          "Not str republican"),
                         independent = c("Ind,near rep",
                                            "Independent",
                                           "Ind,near dem"),
                         democrat = c("Not str democrat",
                                        "Strong democrat"),
                         other = c("No answer",
                                     "Don't know",
                                     "Other party"))) %>% 
  count(year, partyid)

df_clean %>% 
  group_by(year) %>% 
  mutate(total = sum(n),
         prop = n/total) %>% 
  ungroup() %>% 
  ggplot(aes(year, prop, color = partyid)) + 
  geom_line(size = 1) +
  geom_point(aes(shape = partyid, size = 2), show.legend = FALSE)
```

