---
title: "Chapter 10"
author: "Paloma"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size = 3)
library(tidyverse)
library(kableExtra)

mytable <- function(df){
  kable(df) %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"),
                full_width = F,
                font_size = 12)}

library(wesanderson)
```

## Chapter 10: Tibbles

### Creating tibbles

- `as_tibble()`

- `tibble()`

- **non-syntactic names**: a number, with spaces or symbols, use `backticks`.

- `tribble()`: for transposed tibble. column headings start with `~` and entries by commas.

- Tibbles print better than df, 10 rows and only the columns that fit in the screen

- `print(n = 10, width = Inf)` is super useful to see only 10 rows and all columns

- For subsetting tibbles are more strict, they never do partial matching. 

- `tibble::enframe()` converts atomic vectors or lists to one- two-column dataframes. For a list it becomes a nested tibble, `deframe()` converts a two column df to a named vector or list. 

```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
) %>% mytable()
```

```{r}
tribble(
  ~x, ~y, ~z,
  "a",2,5.6,
  "b", 4, 6.89
) %>% mytable()
```

```{r}
enframe(c(a = 5, b = 7)) %>% mytable()
enframe(list(one = 1, two = 2:3, three = 4:6))
deframe(tibble(a = 1:3))
deframe(tibble(a = as.list(1:3)))
```

## Chapter 11: Data import

### Getting started

- `read_csv()` for comma delimited files, `read_csv2()` for semicolon separated files (docs with `,` for decimals)

- `skip = number of rows with metadata` = to skip rows with metadata

- `na = "."` = when missing is treated with symbol, we can specify to take as `NA`.

```{r}
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2) %>% mytable()

```

```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE) %>% mytable()
```

```{r}
read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z")) %>% mytable()

```

```{r}
read_csv("a,b\n1,2,3\n4,5,6")
read_csv("a,b,c\n1,2\n1,2,3,4")
read_csv("a,b\n\"1")
read_csv("a,b\n1,2\na,b")
read_csv("a;b\n1;3")
read_csv2("a;b\n1;3")
```
  
### 3. Parsing a vector

**Numbers**
```{r}
parse_integer(c("123", "345", "abc", "123.45"))

parse_double("1,23", locale = locale(decimal_mark = ","))

parse_number("It cost $123.45")

parse_number("$123,456.789")

parse_number("123.456,789", locale = locale(grouping_mark = "."))
```

**Factors**

```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit)
```

**Date/time**

```{r}
parse_date("04-05-19", "%m-%d-%y")

parse_date("04-jan-19", "%d-%b-%y")

parse_date("04/January/2019", "%d/%B/%Y")

```
