---
title: "Chapter 12"
author: "Paloma"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size = 3)
library(tidyverse)
library(kableExtra)

mytable <- function(df){
  kable(df) %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"),
                full_width = F,
                font_size = 12)}
```

## The rules of Tidy data {.tabset}

1. Each **variable** must have its own **column**

2. Each **observation** must have its own **row**

3. Each **value** must have its own **cell**

### `Pivot_longer()`

- First select the variables

- `names_to` creates the column for the names of the variables. Name is in quoted.

- `values_to` creates the column for the values contained in the columns. Name is in quoted.

- When more there is more than one column that needs to be created, let say repeated measurements for x and y (x1,x2,y1,y2), than we only use `names_to = c(".value", "time")` (so we have a variable "x" and a variable "y"), plus the variable "time" for 1 and 2 

- To detect what will be the name of the variables (x, y) from the piece of information that will stay in the variable "time" (1:2) We need to include as an argument either: 

  + `names_pattern`: similar to `extract`, a regex containing matching groups `()`.
  
  + `names_sep`: as `separate`, can be regex or a numeric vector specifying where to separate.

```{r}
table4a %>% mytable()
```

```{r}
table4a %>% 
  pivot_longer(-1,
               names_to = "year",
               values_to = "cases") %>% 
  mytable()
```

```{r}
table1 %>% mytable()

table1 %>%
  pivot_longer(-c(1,2),
               names_to = "measurement",
               values_to = "count") %>% 
  mytable()
```

```{r}
tidyr::billboard %>%
  select(1:8) %>%
  head() %>%
  mytable()
```

```{r}
billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  ) %>%
  head() %>%
  mytable()
```

```{r}
ex <- tibble(
  id = c(1:7),
  x1 = c(-3:3),
  x2 = c(4:10),
  y1 = c(-4:2),
  y2 = c(rep(2, 4), rep(3,3))
)
```

```{r}
ex %>%
pivot_longer(-1,
             names_to = c(".value", "measure"),
             names_pattern = "(.)(.)")  %>%
  head()  %>%
  mytable()

ex_long <- ex %>%
pivot_longer(-1,
             names_to = c(".value", "measure"),
             names_sep = -1)

ex_long %>% 
  head() %>% 
  mytable()
```

```{r}
who %>% pivot_longer(
  cols = new_sp_m014:newrel_f65,
  names_to = c("diagnosis", "gender", "age"),
  names_pattern = "new_?(.*)_(.)(.*)",
  values_to = "count",
  values_drop_na = TRUE
)
```

#### Examples with matching groups with `names_pattern()` and `"(.)"`.

```{r}
x <- "palo_23r1"

str_view(x, "(.)")
str_view(x, "(.)$")
str_view(x, "(.)(_)")
str_view(x, "(.*)(_)")
str_view(x, "_(.*)")
str_view(x, "(.)(.)$")
str_view(x, "(.*)_(.)(.)")
str_view(x, "(.*)")
str_view(x, "(.*)(.)(.)")


test <- tibble(
  id = c(1:3),
  palo_23_r1 = c(-3:-1),
  palo_23_s2 = c(4:6),
)

test %>% 
  mytable()

test %>% 
  pivot_longer(- id,
    names_to = c("palo"),
    values_to = "count",
    names_pattern = "(.*)") %>% 
  mytable()

test %>% 
  pivot_longer(- id,
    names_to = c("palo", "test", "test2"),
    values_to = "count",
    names_pattern = "(.*)(.)(.)") %>% 
  mytable()

test %>% 
  pivot_longer(- id,
    names_to = c("apellido", "tiempo"),
    values_to = "count",
    names_pattern = "(.)(.)$") %>% 
  mytable()

```

### `Pivot_wider()`

- I don't need to specify which variables will not be modified.

- `names_from` creates the variables from the names column, this is not quoted

- `values_from` creates values from the values column, values column is not quoted.


```{r}
table2 %>% 
  head() %>% 
  mytable()
```

```{r}
table2 %>% 
  pivot_wider(names_from = type,
              values_from = count) %>% 
  mytable()

table2b <- table2 %>% 
  pivot_wider(names_from = c("type", "year"),
              values_from = count) 

table2b %>% 
  mytable()

table2 %>% 
  pivot_wider(names_from = year,
              values_from = count) %>% 
  mytable()

table2 %>% 
  pivot_wider(names_from = year,
              values_from = count) %>% 
  pivot_wider(names_from = type,
              values_from = c(`1999`, `2000`)) %>% 
  mytable()

```

**Going back to longer for table 2b**

```{r}
table2b %>% 
  pivot_longer(-country,
               names_to = c("type", "year"),
               values_to = "count",
               names_sep = "_") %>% 
  mytable()

table2b %>% 
  pivot_longer(-country,
               names_to = c("type", "year"),
               values_to = "count",
               names_pattern = "(.*)_(.*)") %>% mytable()

```

**Going wide for multiple variables**

```{r}
ex_long %>% 
  pivot_wider(
    names_from = measure,
    values_from = c(3,4),
  ) %>% mytable()

ex_long %>% 
  pivot_wider(
    names_from = measure,
    values_from = c(3,4),
    names_sep = ""
  ) %>% mytable()

```

