---
title: "Chap3"
author: "Paloma"
date: "11/28/2019"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Chapter 3

## 3.1 Intro


## 3.2 First steps

The idea is that we have:
`ggplot(data) + <GEOM_FUNCTION(aes(), other feat>`

```{r out.width=c('50%','50%'), fig.show='hold'}
str(mpg)

dim(mpg)

ggplot(data = mpg,
       aes(hwy, cyl)) +
  geom_point()

ggplot(data = mpg,
       aes(class, drv)) +
  geom_point()

```

We cant see the distribution of observations across categories.

## 3.3 Aesthetic mapping

Key tip:
`group`: does not give a legend

`color`: gives a legend, we need to `set.legend = FALSE` if we don't want it.

`size`: not advised for discrete categories, ideally for weights or a count variable

`shape`: is good for categorical, but it only has 25 built in shapes.

`fill`: can also give color to some geom, but not `geom_plot`. I think it is similar to the shape/color and fill relationship.

```{r out.width=c('50%','50%'), fig.show='hold'}
ggplot(data = mpg,
       aes(hwy, cyl, group = class)) +
  geom_point() +
  labs(title = "using group argument")

ggplot(data = mpg,
       aes(hwy, cyl, color = class)) +
  geom_point() +
    labs(title = "using color argument")

ggplot(data = mpg,
       aes(hwy, cyl, shape = class)) +
  geom_point() +
    labs(title = "using shape argument")

ggplot(data = mpg,
       aes(hwy, cyl, fill = class)) +
  geom_point() +
    labs(title = "using fill argument")

ggplot(data = mpg,
       aes(class, hwy)) +
  geom_boxplot(aes(fill = class)) +
    labs(title = "using fill argument")


```

If we want only one shape or color, that would NOT have to be inside the `aes()`, but as an argument of the `<geom_function()>`

About the `shape`, it goes with `color` and `fill`.

- From 0 to 14 *borders* are determined by `color`

- From 15 to 18 are *filled* by `color`

- From 21 - 24 the *border* is defined by `color` and *filled* 
with `fill`

<div class = "row">
<div class = "col-md-4">
```{r}
ggplot(data = mpg, aes(hwy, cyl)) +
  geom_point(shape = 3, color = "orange", size = 5) +
    labs(title = "using shape argument")
```
</div>
<div class = "col-md-4">
```{r}
ggplot(data = mpg, aes(hwy, cyl)) +
  geom_point(shape = 16, color = "orange", size = 5) +
    labs(title = "using shape argument")
```
</div>
<div class = "col-md-4">
```{r}
ggplot(data = mpg, aes(hwy, cyl)) +
  geom_point(shape = 22, color = "orange", fill = "yellow", size = 5) +
    labs(title = "using shape argument")
```
</div>
</div>

## 3.4 Common problems

Just remeber to use `?function_name`.

## 3.5 Facets

`facet_wrap` = makes rows, and usually goes like `facet_wrap(~ cat_var)`

We can add `nrow` and `ncol`, `strip.position` (which is the version for `switch`) changes where the title of each category goes, could be "right", "left", "bottom", etc..

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 4, ncol = 2, strip.position = "bottom")
```


- `facet_grid` = Makes columns, it usually goes like `facet_grid(.~ cat_var)`

<div class = "row">
<div class = "col-md-6">
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ class)
```
</div>
<div class = "col-md-6">
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ class)
```
</div>
</div>



## 3.6 Geometric objects

The key concept is that not every geom has the same aesthetics.

For example `geom_smooth` has the "linetype" argument that shows a different line for each `cat_var`.

If I had used the `group` instead of color, I would have had three lines from the same color, and no legend.

```{r}
mpg %>% 
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(aes(linetype = drv), show.legend = FALSE) +
  geom_point(show.legend = FALSE, position = "jitter")
```

The idea is that for each geom, the aes apply to each layer.

```{r}
mpg %>% 
  ggplot(aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class), show.legend = TRUE) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = TRUE) +
  geom_smooth(data = filter(mpg, class == "suv"), se = TRUE)
  
```

A line chart?

```{r}
mpg %>% 
  ggplot(aes(x = displ, y = hwy, color =  class)) + 
  geom_line() +
  facet_wrap(. ~ class)

```

Some graphs

<div class = "row">
<div class = "col-md-6">
```{r}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point(size = 3) +
  geom_smooth(se = FALSE, size = 3)
```
</div>
<div class = "col-md-6">
```{r}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_smooth(aes(group =  drv), se = FALSE, size = 2 ) +
  geom_point(size = 3)
```
</div>
</div>

<div class = "row">
<div class = "col-md-6">
```{r}
mpg %>% 
  ggplot(aes(displ, hwy, color = drv)) +
  geom_smooth(se = FALSE, size = 2 ) +
  geom_point(size = 3)
```
</div>
</div class = "col-md-6">
```{r}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point(aes(color = drv),size = 3, show.legend = FALSE) +
  geom_smooth(se = FALSE, size = 2 )
```
</div>
</div>


## 3.7 Statistical transformations

Geom_bar and stat_count are very similar. By default `geom_bar` has as an argument `stat = "count"`, that if we change by "identity", is going to look different.

</div class = "row">
</div class = "col-md-3">
```{r}
diamonds %>% 
  ggplot(aes(x = cut)) +
  geom_bar(aes(y = ..prop.., group = 1), show.legend = FALSE)
```
</div>
<div class = "col-md-3">
```{r}
diamonds %>% 
  ggplot(aes(x = cut, fill = clarity)) +
  geom_bar()
```
</div>
<div class = "col-md-3">
```{r}
diamonds %>%
  count(cut) %>% 
  ggplot(aes(x = cut, y  = n)) +
  geom_bar(stat = "identity")
```
</div>
<div class = "col-md-3">
```{r}
diamonds %>% 
  ggplot(aes(x = cut)) +
  stat_count()
```
</div>
</div>


Now `stat_summary` could save me lots of trouble

Please notice that stat_summary has `show.legend`

```{r}
diamonds %>%
  ggplot() +
  stat_summary(
    aes(x = cut, y = depth, color = cut),
    fun.y = median,
    fun.ymax = max,
    fun.ymin = min,
    show.legend = FALSE) +
  coord_flip()
```

`fun.data` crea un df con la media y ds
Otra para revisar que esta buena es esta para lineranges, que seria trazar lineas y puntos con la min y max, se, etc.
https://ggplot2.tidyverse.org/reference/geom_linerange.html


## 3.8 Position adjustments
In barcharts, if we use in the `fill` (or `color`) argument a different `cat_var`, we get a "stacked" look.

We can change this with the argument `position` that use 3 options: "identity", "dodge", "fill".

</div class = "row">
</div class = "col-md-3">
```{r}
diamonds %>% 
  ggplot(aes(x = cut, fill = clarity)) +
  geom_bar()
```
</div>
</div class = "col-md-3">
```{r}
diamonds %>% 
  ggplot(aes(x = cut, color = clarity)) +
  geom_bar(fill = NA, position = "identity")
```
</div>
</div class = "col-md-3">
```{r}
diamonds %>% 
  ggplot(aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill")
```
</div>
</div class = "col-md-3">
```{r}
diamonds %>% 
  ggplot(aes(x = cut, fill = clarity)) +
  geom_bar(position = "dodge")
```
</div>
</div>


Exercises have to do with geom_jitter

```{r out.width=c('25%','25%','25%','25%'), fig.show='hold'}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(position = "jitter")

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_count()

ggplot(data = mpg, mapping = aes(x = class, y =  cty)) + 
  geom_boxplot(aes(fill = class)) 
```

## 3.9 Coordinate systems

`coord_polar`

```{r}
diamonds %>% 
  ggplot(aes(x = cut, fill = cut)) +
  geom_bar(width = 1, show.legend = FALSE) +
  coord_polar() +
  theme(aspect.ratio = 1)

```

## 3.10 The layered grammar of graphics


```{r eval = FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```


