---
title: "Chapter 12"
author: "Paloma"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size = 3)
library(tidyverse)
library(kableExtra)

mytable <- function(df){
  kable(df) %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"),
                full_width = F,
                font_size = 12)}
```

## The rules of Tidy data {.tabset}

1. Each **variable** must have its own **column**

2. Each **observation** must have its own **row**

3. Each **value** must have its own **cell**

### `Pivot_longer()`

- First select the variables

- `names_to` creates the column for the names of the variables. Name is in quoted.

- `values_to` creates the column for the values contained in the columns. Name is in quoted.

- When more there is more than one column that needs to be created, let say repeated measurements for x and y (x1,x2,y1,y2), than we only use `names_to = c(".value", "time")` (so we have a variable "x" and a variable "y"), plus the variable "time" for 1 and 2 

- To detect what will be the name of the variables (x, y) from the piece of information that will stay in the variable "time" (1:2) We need to include as an argument either: 

  + `names_pattern`: similar to `extract`, a regex containing matching groups `()`.
  
  + `names_sep`: as `separate`, can be regex or a numeric vector specifying where to separate.

```{r}
table4a %>% mytable()
```

```{r}
table4a %>% 
  pivot_longer(-1,
               names_to = "year",
               values_to = "cases") %>% 
  mytable()
```

```{r}
table1 %>% mytable()

table1 %>%
  pivot_longer(-c(1,2),
               names_to = "measurement",
               values_to = "count") %>% 
  mytable()
```

```{r}
tidyr::billboard %>%
  select(1:8) %>%
  head() %>%
  mytable()
```

```{r}
billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  ) %>%
  head() %>%
  mytable()
```

```{r}
ex <- tibble(
  id = c(1:7),
  x1 = c(-3:3),
  x2 = c(4:10),
  y1 = c(-4:2),
  y2 = c(rep(2, 4), rep(3,3))
)
```

```{r}
ex %>%
pivot_longer(-1,
             names_to = c(".value", "measure"),
             names_pattern = "(.)(.)")  %>%
  head()  %>%
  mytable()

ex_long <- ex %>%
pivot_longer(-1,
             names_to = c(".value", "measure"),
             names_sep = -1)

ex_long %>% 
  head() %>% 
  mytable()
```

```{r}
who %>% pivot_longer(
  cols = new_sp_m014:newrel_f65,
  names_to = c("diagnosis", "gender", "age"),
  names_pattern = "new_?(.*)_(.)(.*)",
  values_to = "count",
  values_drop_na = TRUE
)
```

#### Examples with matching groups with `names_pattern()` and `"(.)"`.

```{r}
x <- "palo_23r1"

str_view(x, "(.)")
str_view(x, "(.)$")
str_view(x, "(.)(_)")
str_view(x, "(.*)(_)")
str_view(x, "_(.*)")
str_view(x, "(.)(.)$")
str_view(x, "(.*)_(.)(.)")
str_view(x, "(.*)")
str_view(x, "(.*)(.)(.)")


test <- tibble(
  id = c(1:3),
  palo_23_r1 = c(-3:-1),
  palo_23_s2 = c(4:6),
)

test %>% 
  mytable()

test %>% 
  pivot_longer(- id,
    names_to = c("palo"),
    values_to = "count",
    names_pattern = "(.*)") %>% 
  mytable()

test %>% 
  pivot_longer(- id,
    names_to = c("palo", "test", "test2"),
    values_to = "count",
    names_pattern = "(.*)(.)(.)") %>% 
  mytable()

test %>% 
  pivot_longer(- id,
    names_to = c("apellido", "tiempo"),
    values_to = "count",
    names_pattern = "(.)(.)$") %>% 
  mytable()

test %>% 
  pivot_longer(- id,
    names_to = c("apellido", "tiempo", "test", "test2"),
    values_to = "count",
    names_pattern = "([a-z]+)_([0-9]+)_(.)(.)") %>% 
  mytable()
```

### `Pivot_wider()`

- I don't need to specify which variables will not be modified.

- `names_from` creates the variables from the names column, this is not quoted

- `values_from` creates values from the values column, values column is not quoted.


```{r}
table2 %>% 
  head() %>% 
  mytable()
```

```{r}
table2 %>% 
  pivot_wider(names_from = type,
              values_from = count) %>% 
  mytable()

table2b <- table2 %>% 
  pivot_wider(names_from = c("type", "year"),
              values_from = count) 

table2b %>% 
  mytable()

table2 %>% 
  pivot_wider(names_from = year,
              values_from = count) %>% 
  mytable()

table2 %>% 
  pivot_wider(names_from = year,
              values_from = count) %>% 
  pivot_wider(names_from = type,
              values_from = c(`1999`, `2000`)) %>% 
  mytable()

```

**Going back to longer for table 2b**

```{r}
table2b %>% 
  pivot_longer(-country,
               names_to = c("type", "year"),
               values_to = "count",
               names_sep = "_") %>% 
  mytable()

table2b %>% 
  pivot_longer(-country,
               names_to = c("type", "year"),
               values_to = "count",
               names_pattern = "(.*)_(.*)") %>% mytable()

```

**Going wide for multiple variables**

```{r}
ex_long %>% 
  pivot_wider(
    names_from = measure,
    values_from = c(3,4),
  ) %>% mytable()

ex_long %>% 
  pivot_wider(
    names_from = measure,
    values_from = c(3,4),
    names_sep = ""
  ) %>% mytable()

```


## Other useful functions from `tidyr()` {.tabset}

### Separate

We use the argument `into` and include the new columns in a vector, with quoted names.

```{r}
table3 %>% 
  mytable()

table3 %>%
  separate(rate, into = c("cases", "population")) %>% 
  mytable()
```

```{r}
stocks <- tibble(
  year   = c("year1", "year2", "year1", "year2"),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
  )
stocks %>% mytable()

stocks %>% 
  separate(year, into = c("text", "time"),
           sep = -1,
           remove = FALSE) %>% 
  mytable()

stocks %>% 
  separate(year, into = c(NA, "time"),
           sep = -1) %>% 
  mytable()
```

```{r}
df <- data.frame(x = c(NA, "a?b", "a.d", "b:c"))

df %>% separate(x, c("A","B"), sep = "([.?:])")
```


```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three") )

tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")

tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "drop")

```


```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "left")

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "right")
```

### Extract

```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  extract(x, into = c("one","two", "three"),
          regex = "(.),(.),(.)")
```

```{r}
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z]+)_([0-9]+)")
```

```{r}
tibble(x = c("X1", "X20", "AA11", "AA2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z]+)([0-9]+)")
```

### Unite

We only need to name the new column, unquoted, followed by the variables to unite, unquoted as well.

```{r}
table1 %>% 
  unite(rate, cases, population,
        sep = "/",
        remove = FALSE)
```

### Missing values

- **Implicit**: Not present in the data

- **Explicit**: As `NA`

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

stocks %>% mytable()
```

```{r}
stocks_wide <- stocks %>% 
  pivot_wider(names_from = year,
              values_from = return) 

stocks_wide %>% 
  mytable() 
```

```{r}
stock_long <- stocks_wide %>% 
  pivot_longer(-1,
               names_to = "year",
               values_to = "return") 

stock_long %>% 
  mytable()
```

#### Complete

Makes **implicit** values **explicit**

```{r}
stocks %>% 
  complete(qtr, year)
```

```{r}
df <- tibble(
  group = c(1:2, 1),
  item_id = c(1:2, 2),
  item_name = c("a", "b", "b"),
  value1 = 1:3,
  value2 = 4:6
)

df %>% mytable()

df %>% 
  complete(group, item_id) %>% mytable()

df %>% 
  complete(group, item_id, item_name) %>% mytable()

df %>% 
  complete(group, nesting(item_id, item_name)) %>% mytable()

df %>% 
  complete(group, nesting(item_id, item_name), fill = list(value1 = 0)) %>% mytable()

df %>% 
  complete(group, nesting(item_id, item_name)) %>% 
  fill(c("value1", "value2"), .direction = "up" ) %>% 
  mytable()

```

## Exercises
```{r}
who_tidy <- who %>%
  pivot_longer(new_sp_m014:newrel_f65,
               names_to = "key",
               values_to = "value",
               values_drop_na = TRUE) %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel")) %>%
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

For each country, year, and sex compute the total number of cases of TB. Make an informative visualisation of the data.

```{r, fig.width= 8}

who_tidy%>% 
  filter(year >= 1995) %>% 
  group_by(country, year, sex) %>% 
  summarise(count = sum(value)) %>% 
  ungroup() %>% 
  group_by(country) %>% 
  mutate(total = sum(count),
         prop = (count / total)*100) %>% 
  ungroup() %>% 
  filter(between(total, quantile(total, 0.95), quantile(total, 1))) %>% 
  mutate(country = fct_rev(fct_reorder(country, total))) %>%
  mutate(count = ifelse(sex == "f", count* -1, count)) %>% 
  mutate(prop = ifelse(sex == "f", prop* -1, prop)) %>% 
  ggplot(aes(country, count, fill = as_factor(year))) +
  geom_col() +
  ggpol::facet_share(~sex, reverse_num = TRUE, scales = "free") +
  coord_flip() +
  theme_minimal()
```

```{r}
who_tidy%>% 
  group_by(country, year, sex) %>% 
  summarise(count = sum(value)) %>% 
  ungroup() %>% 
  filter(year >= 1995) %>% 
  ggplot(aes(year, log(count), color = country)) +
  geom_line(show.legend = FALSE) +
  geom_smooth(aes(color = sex), show.legend = FALSE) +
  facet_grid(~sex)
```

