---
title: "Chapter 13"
author: "Paloma"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size = 3)
library(tidyverse)
library(kableExtra)
library(nycflights13)

mytable <- function(df){
  kable(df) %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"),
                full_width = F,
                font_size = 12)}
```

## Intro:

Functions for working with relational data are divided in 3 groups:

- `Mutating joins`: adds new variables from one df from matching observations in another

- `Filtering joins`: filters observations from one df based on whether they match on observations from other table

- `Set operations`: treats observations as set elements

- A **primary key** uniquely identifies an observation in its own table.

- A **foreign key** uniquely identifies an observation in another table.

- Once the **primary key** is identified, it is good practice to see if they uniquely identify each observation.

- A **surrogate key**, a key created when a primary key is absent, with `mutate` and `row_number()`.

## Mutating Joins
```{r}
x <- tibble(id = c(1:3),
            val_x = c("banana", "peach", "pinneaple"))

y <- tibble(id = c(1,2,4),
            val_y = c("chocolate", "mint", "vainilla"))
```

**Inner join**: All rows from x that have matching values in y, and all columns from x and y
```{r}
x %>% 
  inner_join(y, by = "id") %>% mytable()
```

**Left join**: All rows from x and all columns from x and y
```{r}
x %>% 
  left_join(y, by = "id") %>% mytable()
```

**Right join**: All rows from y and all columns from x and y
```{r}
x %>% 
  right_join(y, by = "id") %>% mytable()
```

**Full join**: All rows from x and y, all columns from x and y
```{r}
x %>% 
  full_join(y, by = "id") %>% mytable()
```

## Filtering joins

Affect **only** observations, not the variables.

**Semi join**: filters rows in x that match in y, but keeps columns from x
```{r}
x %>% 
  semi_join(y, by = "id") %>% mytable()

y %>% 
  semi_join(x, by = "id") %>% mytable()
```

**Anti join**: filters rows in x and y that have no match, but keeps columns from x
```{r}
x %>% 
  anti_join(y, by = "id") %>% mytable()

y %>% 
  anti_join(x, by = "id") %>% mytable()
```

## Set operations

It expects that X and Y have the same variables

**Union**: All unique rows from x and y

```{r}
x <- tibble(id = c(1:3),
            x = c("banana", "peach", "pinapple"))

y <- tibble(id = c(1,2,4),
            x = c("banana", "tangerine", "apple"))

x %>% union(y) %>% mytable()
```

**Union all**: All rows from x and y, keeps duplicates

```{r}
x %>% 
  union_all(y) %>% mytable()

y %>% 
  union_all(x) %>% mytable()
```

**Set Difference**: All rows from x that are not rows from y, keeping just unique rows.

```{r}
x %>% 
  setdiff(y) %>% mytable()

y %>% 
  setdiff(x) %>% mytable()
```

**Intersect**: All observations that are the same for x and y

```{r}
x %>% 
  intersect(y) %>% mytable()

y %>% 
  intersect(x) %>% mytable() 
```

## Extra

This link is the best: https://www.garrickadenbuie.com/project/tidyexplain/
